Microlensing as a possible probe of event-horizon structure in quasars

### 2017-11-26 paper1/ms.tex
checking...

\subsection{Lightcurve of the crescent shaped source}

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of a
  simulated source image is presented in the upper panel. The source
  image was presented in Dexter et al (2012) in the top right-most
  panel of Figure~5 (of the respective paper) as a result of the
  simulations presented in McKinney \& Blandford (2009). The
  corresponding lightcurve associated to the 1D profile is presented
  in the lower panel. The points $\rm p_2$ and $\rm p_3$ mark the
  start and end of the equivalent dark disk and are associated with
  the moments of their overlap with the caustic: $t_2$ and $t_3$;
  $h_{w}$ denotes the maximum width of the crescent.}
\end{figure}

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Do do so, we first generate a microlensing lightcurve
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources.  We then fit model or template lightcurves ---which
are generated with varied source and other parameters--- to the mock
lightcurve.  The basic technique is the same as in previous studies
\citep[e.g.,][]{2010ApJ...712..658P} though the details are different.

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} Mock lightcurve fitted with the
  correct source type and caustic.  The upper panel shows the mock
  data points, and running through them is a seven-parameter best fit:
  these being the crescent parameters $(R_p,R_n,a,b)$ and additionally
  the start and end times and the overall normalisation.  (Ideal
  lightcurves for the same crescent parameters appear above in
  Figures~\ref{fig:lightcurve_gauss}, \ref{fig:a_var},
  \ref{fig:lightcurve_crescent} and \ref{fig:char_points}.)  The curve
  has $\chi^2_{(244)}=243, P=0.32$ indicating a good fit.  (The
  subscript 244 refers to degrees of freedom: 251~points less
  7~parameters.)  The lower panel shows the posterior distributions
  for $r_{1/2}/v_p$ (solid histogram) and $R_n/R_p$ (dashed histogram).
  Vertical dashes mark the correct values.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} Like Figure~\ref{fig:cc_forward},
  except that the fit is derived from a slightly different caustic
  than the one used to generate the mock data.  The curve has
  $\chi^2_{(244)}=284, P=0.01$ indicating a marginally acceptable
  fit.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} Like Figures~\ref{fig:cc_forward} and
  \ref{fig:cc_backward}, but showing the best fit to a Gaussian
  source.  The histogram shows the posterior distribution of $r_{1/2}/v_p$
  (there is no $R_n$).  It appears to show $r_{1/2}/v_p$ as accurately
  recovered, but is not to be trusted, because the fit has
  $\chi^2_{(247)}=344, P=3\times10^{-6}$ indicating rejection of the
  Gaussian source.  Looking closely, the curve is a little too shallow
  near $t=3$, and it is unable to fit the second sub-peak near $t=5$.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{compare-mag.eps}
\caption{[{\bf Temporary.} This figure is included only the in
    red/blue version, and is not for the paper.] Overlay from the
  three preceding figures.}
\end{figure}

@@

The mock data appear in Figures
\ref{fig:cc_forward}--\ref{fig:gc_forward}.  Illustrative values have
been adopted for the frequency of observations and for the noise
level.  It is, of course, important is that the observations have
enough sampling and signal-to-noise that the characteristic features
of a crescent source (see e.g., Figure~\ref{fig:char_points}) are not
washed out.

@@

Figure \ref{fig:cc_backward} attempts a somewhat more realistic
situation.  The mock lightcurve is the same as before, but the
templates come from taking a crescent source along the path AB in
Figure~\ref{fig:magnification_map}.  That is, the mock data and the
templates come from crossings of different caustics.  This is intended
to mimic the systematic errors arising from not knowing the caustic
exactly.  As Figure~\ref{fig:cc_backward} shows, the fit is slightly
degraded, but still acceptable.  The parameters $r_{1/2}/v_p$ and
$R_n/R_p$ are still recovered.  The parameters $a/R_p$ and $b/R_p$
(not shown in the Figure) giving the orientation of the crescent were
not correctly recovered; but this expected, because the AB and CB
tracks have different orientations with respect to the caustics
crossed.  The slight degradation of the fit can be understood as
follows.  The source track will not, in general, be normal to the
caustic, but will instead have a non-normal velocity component.  If
the caustic is perfectly straight, the velocity component along the
caustic makes no difference.  (Hence our model only considers only the
normal velocity component $v_p$).  But in fact the caustics, although
clean folds do have some curvature.  The combination of a curving
caustic and non-normal velocity of the source makes the orientation of
the source twist with respect to the caustic.  This twist would be
different for crossings along the AB and CB tracks.

@@

In the current paper, we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of the flux of light from the source in the proximity of a fold caustic, 
we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal approximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated by
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
%In particular cases in which the previously mentioned approximation loses its validity, the impact on the quality of the 
%lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
%The data points corresponding to the source position before and during the overlapping of the caustic will be affected by
% smaller errors than the data points corresponding to later times. \\

@@

Nevertheless, the respective displacement is relevant for the calculation of
of the half-light radius.  Its qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disc ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.

@@

With the different source profiles and their corresponding lightcurves 
studied we can change our point of view of the system to that of an observer. 
The observer would basically detect only the lightcurve of such a source.
 As described in section 4.3 the timing of the onset and 
offset of the previously 
described periods can be used to estimate the values of the radii 
and one of the displacement parameters when assuming a geometric crescent. 
All quantities can be estimated in terms of the relative velocity 
of the source in a direction perpendicular to the caustic.

@@

In the case of a high-quality lightcurve, the parameters of a crescent
source can be estimated by simply identifying signatures on the
lightcurve corresponding to transits of the particular source features
across the caustic.  This simple idea can be supplemented with
statistics.  We do so in Section~\ref{sec:numerics}, where we take a
realistic magnification map generated by the microlensing code
from \cite{1999A&A...346L...5W}, compute a caustic-crossing lightcurve
from it, and then fit the lightcurve with other caustic-crossing
templates.  The conclusion is that a crescent source can be
distinguised from a circular source, on the basis of caustic-crossing
lightcurves, if the caustic is a clean but not ideal fold.

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disc would
provide an estimate of the size of the accretion disc and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic, while $v_p$
denotes the component of the relative velocity of the source and fold
which is perpendicular to the caustic.  The respective velocity is an
unknown, though it can be constrained on a case by case basis to an
order of magnitude or even better. This would require the study of the
dynamics of the stellar structure which contains the gravitational
lens and the bulk transverse velocity of the galaxy with respect to
the observer.  A better estimate of the relative velocity would
facilitate a better estimate of the effective non-rotating black hole
mass associated with the black hole shadow.

>>>>

\subsection{Lightcurve of the crescent shaped source}\label{subsec:lcurvcres}

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordinate axis of a
  simulated source image is presented in the upper panel. The source
  image was presented in Dexter et al (2012) in the top right-most
  panel of Figure~5 (of the respective paper) as a result of the
  simulations presented in McKinney \& Blandford (2009). The
  corresponding lightcurve associated to the 1D profile is presented
  in the lower panel. The points $\rm p_2$ and $\rm p_3$ mark the
  start and end of the equivalent dark disk and are associated with
  the moments of their overlap with the caustic: $t_2$ and $t_3$;
  $h_{w}$ denotes the maximum width of the crescent.}
\end{figure}

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  To do so, we first generate a microlensing lightcurve
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources.  We then fit model or template lightcurves ---which
are generated with varied source and other parameters--- to the mock
lightcurve.  The basic technique is the same as in previous studies
\citep[e.g.,][]{2010ApJ...712..658P} though the details are different.

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} Mock lightcurve fitted with the
  correct source type and caustic.  The upper panel shows the mock
  data points, and running through them is a seven-parameter best fit:
  these being the crescent parameters $(R_p,R_n,a,b)$ and additionally
  the start and end times and the overall normalisation.  (Ideal
  lightcurves for the same crescent parameters appear above in
  Figures~\ref{fig:lightcurve_gauss}, \ref{fig:a_var},
  \ref{fig:lightcurve_crescent} and \ref{fig:char_points}.) The curve
  has $\chi^2_{(244)}=243, P=0.32$ indicating a good fit.  (The
  subscript 244 refers to degrees of freedom: 251~points less
  7~parameters.)  The lower panel shows the posterior distributions
  for $r_{1/2}/v_p$ (solid histogram) and $R_n/R_p$ (dashed
  histogram).  Vertical dashes mark the correct values.  Inferred are
  $r_{1/2}/v_p=0.90^{+0.15}_{-.07}$ and $R_n/R_p=0.32^{+0.19}_{-0.08}$
  (90\% confidence).}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} Like Figure~\ref{fig:cc_forward},
  except that the fit is derived from a slightly different caustic
  than the one used to generate the mock data.  The curve has
  $\chi^2_{(244)}=284, P=0.01$ indicating a marginally acceptable
  fit. Inferred are $r_{1/2}/v_p=0.92^{+0.06}_{-0.04}$ and
  $R_n/R_p=0.32^{+0.12}_{-0.07}$ (90\% confidence).}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} Like Figures~\ref{fig:cc_forward} and
  \ref{fig:cc_backward}, but showing the best fit to a Gaussian
  source.  The histogram shows the posterior distribution of $r_{1/2}/v_p$
  (there is no $R_n$).  It appears to show $r_{1/2}/v_p$ as accurately
  recovered, but is not to be trusted, because the fit has
  $\chi^2_{(247)}=344, P=3\times10^{-6}$ indicating rejection of the
  Gaussian source.  Looking closely, the curve is a little too shallow
  near $t=3$, and it is unable to fit the second sub-peak near $t=5$.
  Inferred is $r_{1/2}/v_p=0.91^{+0.05}_{-0.01}$ (90\% confidence).}
\end{figure}

@@

The mock data appear in Figures
\ref{fig:cc_forward}--\ref{fig:gc_forward}.  Illustrative values have
been adopted for the frequency of observations and for the noise
level.  It is, of course, important that the observations have
enough sampling and signal-to-noise that the characteristic features
of a crescent source (see e.g., Figure~\ref{fig:char_points}) are not
washed out.

@@

Figure \ref{fig:cc_backward} attempts a somewhat more realistic
situation.  The mock lightcurve is the same as before, but the
templates come from taking a crescent source along the path AB in
Figure~\ref{fig:magnification_map}.  That is, the mock data and the
templates come from crossings of different caustics.  This is intended
to mimic the systematic errors arising from not knowing the caustic
exactly.  As Figure~\ref{fig:cc_backward} shows, the fit is slightly
degraded, but still acceptable.  The parameters $r_{1/2}/v_p$ and
$R_n/R_p$ are still recovered, though less well; the correct
value of $r_{1/2}/v_p$ lies outside 90\% of the MCMC distribution,
but within 99\%.   The parameters $a/R_p$ and $b/R_p$
(not shown in the Figure) giving the orientation of the crescent were
not correctly recovered; but this expected, because the AB and CB
tracks have different orientations with respect to the caustics
crossed.  The slight degradation of the fit can be understood as
follows.  The source track will not, in general, be normal to the
caustic, but will instead have a non-normal velocity component.  If
the caustic is perfectly straight, the velocity component along the
caustic makes no difference.  (Hence our model only considers only the
normal velocity component $v_p$).  But in fact the caustics, although
clean folds do have some curvature.  The combination of a curving
caustic and non-normal velocity of the source makes the orientation of
the source twist with respect to the caustic.  This twist would be
different for crossings along the AB and CB tracks.

@@

In the current paper, we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of the flux of light from the source in the proximity of a fold caustic, 
we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal approximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated by
 a line due to reason that the local radius of curvature of the caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
%In particular cases in which the previously mentioned approximation loses its validity, the impact on the quality of the 
%lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
%The data points corresponding to the source position before and during the overlapping of the caustic will be affected by
% smaller errors than the data points corresponding to later times. \\

@@

Nevertheless, the respective displacement is relevant for the calculation
of the half-light radius.  Its qualitative effect is to 
decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  In contrast to the simpler source shapes discussed before the
crescent has two additional relevant parameters: the radius of the dark 
disc ($R_n$) and the center displacement in the direction
 perpendicular to the caustic $a$.

@@

With the different source profiles and their corresponding lightcurves 
studied we can change our point of view of the system to that of an observer. 
The observer would basically detect only the lightcurve of such a source.
 As described in section~\ref{subsec:lcurvcres} the timing of the onset and 
offset of the previously 
described periods can be used to estimate the values of the radii 
and one of the displacement parameters when assuming a geometric crescent. 
All quantities can be estimated in terms of the relative velocity 
of the source in a direction perpendicular to the caustic.

@@

In the case of a high-quality lightcurve, provided the caustic is a
fairly clean fold, or equivalently, the source is much smaller than
the local radius of curvature of the caustic, the crossing times of 
the bright and dark areas (cf.~Figure~\ref{fig:char_points}) could
be estimated by simply identifying signatures on the
lightcurve corresponding to transits of the particular source features
across the caustic.  This simple idea can be supplemented with
statistics.  In Section~\ref{sec:numerics}, we take a
realistic magnification map generated by the microlensing code
from \cite{1999A&A...346L...5W}, compute a caustic-crossing lightcurve
from it, and then fit the lightcurve with other caustic-crossing
templates.  The conclusion is that a crescent source can be
distinguised from a circular source, on the basis of caustic-crossing
lightcurves, if the caustic is a clean but not ideal fold.

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disc would
provide an estimate of the size of the accretion disc and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic, while $v_p$
denotes the component of the relative velocity of the source and fold
which is perpendicular to the caustic.

% The respective velocity is an
% unknown, though it can be constrained on a case by case basis to an
% order of magnitude or even better. This would require the study of the
% dynamics of the stellar structure which contains the gravitational
% lens and the bulk transverse velocity of the galaxy with respect to
% the observer.  A better estimate of the relative velocity would
% facilitate a better estimate of the effective non-rotating black hole
% mass associated with the black hole shadow.
